```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Data Access Instructions
The data used in this project is located in a folder structure as follows:
-> archive/ipl_match_info_data.csv // kaggle complete ipl dataset by BISWAJIT BRAHMMA


-> archive/ipl_match_ball_by_ball_data.csv // kaggle complete ipl dataset by BISWAJIT BRAHMMA


-> IPL - Player Performance Dataset/All Seasons Combined/*.csv // kaggle dataset by SOURAV BANERJE


-> player_team_map_fully_filled.csv // mapped manually 

The dataset used in this project is too large to store in the GitHub repository. You can download it from the following [Google Drive link](https://drive.google.com/drive/folders/1YxPej6-Ezt56-W0s_fwWssnFEGaHqCg4?usp=sharing).



1 : libraries
      This project uses tidyverse, data.table, and lubridate for efficient data manipulation and date handling. caret, xgboost, and randomForest support machine learning tasks. For building interactive dashboards, shiny, shinydashboard, DT, and shinyWidgets enhance app functionality and user interface.
```{r}
library(tidyverse)
library(data.table)
library(lubridate)
library(caret)
library(xgboost)
library(randomForest)
library(shiny)
library(shinydashboard)
library(DT)
library(shinyWidgets)
```

2 : Datasets
      We used multiple CSV datasets extracted from Kaggle and other sources. These included:
- Ball-by-ball match data
- Match summary data
- Player performance datasets (runs, wickets, sixes, centuries, etc.)

We used `fread()` functions to load these datasets. 
```{r}

setwd("/Users/hareena/R project adv")  

# Load ball-by-ball and match info data
ball_data <- fread("archive/ipl_match_ball_by_ball_data.csv")
match_info <- fread("archive/ipl_match_info_data.csv")

# Load player performance datasets
performance_path <- "IPL - Player Performance Dataset/All Seasons Combined"

# Utility function to read and clean each file
safe_fread <- function(file) {
  df <- fread(file)
  names(df)[names(df) == ""] <- paste0("unnamed_col_", which(names(df) == ""))
  names(df) <- make.names(names(df), unique = TRUE)
  return(df)
}

# All performance files into a named list
performance_files <- list.files(performance_path, pattern = "*.csv", full.names = TRUE)
names(performance_files) <- gsub(".csv", "", basename(performance_files))

# Load and clean using safe_fread
performance_data <- lapply(performance_files, safe_fread)

```

3 : Preprocessing and Wrangling
      All data was cleaned using make.names() to standardize column names, and duplicates/missing values were handled accordingly. We merged datasets using common keys like match_id and standardized the date formats using lubridate::ymd().
      
```{r}
# Cleaning column names
library(data.table)
library(dplyr)
library(lubridate)

# Clean column names for primary datasets
names(ball_data) <- make.names(names(ball_data), unique = TRUE)
names(match_info) <- make.names(names(match_info), unique = TRUE)

# Merging and preprocessing match data
combined_data <- merge(ball_data, match_info, by = "match_id")
combined_data$date <- ymd(combined_data$date)
match_info$date <- ymd(match_info$date)
match_info <- match_info[!is.na(winner) & result_type != "no result"]

# Loading and cleaning player performance datasets
performance_path <- "IPL - Player Performance Dataset/All Seasons Combined"
performance_files <- list.files(performance_path, pattern = "*.csv", full.names = TRUE)
names(performance_files) <- gsub(".csv", "", basename(performance_files))

# Safe loader with empty name fix
performance_data <- lapply(performance_files, function(file) {
  df <- fread(file)
  
  # Handle empty column names explicitly
  names(df)[names(df) == ""] <- paste0("unnamed_col_", which(names(df) == ""))
  names(df) <- make.names(names(df), unique = TRUE)
  
  # Clean and coerce types
  df <- df %>%
    distinct() %>%
    mutate(across(where(is.character), ~na_if(., ""))) %>%
    mutate(across(where(is.character), \(x) type.convert(x, as.is = TRUE)))
  
  return(df)
})


```

4 : Exploratory Data analysis (EDA)
    We explored the data using various plots:
    
### Match-level Insights
- Total Wins by Team: Bar plots showing win counts revealed historically dominant teams.
- Toss Decision Trends: Visualized how often teams choose to bat or bowl.
- Toss vs Match Outcome: Analyzed if winning the toss gave an advantage.
```{r}
library(ggplot2)
library(dplyr)
library(forcats)

# Wins by team
match_info %>%
  filter(!is.na(winner)) %>%
  count(winner, sort = TRUE) %>%
  ggplot(aes(x = fct_reorder(winner, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Total Wins by Team", x = "Team", y = "Wins")

# Toss vs match win
match_info %>%
  mutate(toss_win_is_win = toss_winner == winner) %>%
  count(toss_win_is_win) %>%
  ggplot(aes(x = toss_win_is_win, y = n, fill = toss_win_is_win)) +
  geom_col() +
  labs(title = "Impact of Toss on Winning", x = "Toss Win = Match Win", y = "Count")

# Toss decision trends
match_info %>%
  count(toss_decision) %>%
  ggplot(aes(x = toss_decision, y = n, fill = toss_decision)) +
  geom_col() +
  labs(title = "Toss Decision Distribution", x = "Decision", y = "Count")
```

### Ball-by-ball Analysis
- Top Run Scorers: Aggregated player scores to identify top batsmen.
- Extras Over Time: Plotted extras (no balls, wides, etc.) year-wise.
```{r}
library(dplyr)
# Top run scorers in dataset
combined_data %>%
  filter(!is.na(striker)) %>%
  group_by(striker) %>%
  summarise(total_runs = sum(runs_off_bat, na.rm = TRUE)) %>%
  arrange(desc(total_runs)) %>%
  head(10) %>%
  ggplot(aes(x = reorder(striker, total_runs), y = total_runs)) +
  geom_col(fill = "tomato") +
  coord_flip() +
  labs(title = "Top 10 Batsmen by Total Runs", x = "Batsman", y = "Runs")

# Extras over time
combined_data %>%
  mutate(
    year = lubridate::year(date),
    extras = coalesce(wides, 0) + coalesce(noballs, 0) + coalesce(byes, 0) + coalesce(legbyes, 0) + coalesce(penalty, 0)
  ) %>%
  group_by(year) %>%
  summarise(total_extras = sum(extras, na.rm = TRUE)) %>%
  ggplot(aes(x = year, y = total_extras)) +
  geom_line(color = "purple", linewidth = 1.2)
  labs(title = "Total Extras by Year", x = "Year", y = "Extras")
```

### Player Stats
- Most Wickets: Top 10 players by wicket count across all seasons.
- Most Sixes in an Innings: Highlighted explosive innings performances.
```{r}
most_wickets <- performance_data[["Most Wickets All Seasons Combine"]]

most_wickets %>%
  group_by(Player) %>%
  summarise(Wkts = sum(Wkts, na.rm = TRUE)) %>%           
  ungroup() %>%
  slice_max(Wkts, n = 10) %>%
  arrange(desc(Wkts)) %>%
  mutate(Player = factor(Player, levels = unique(Player))) %>%
  ggplot(aes(x = Player, y = Wkts)) +
  geom_col(fill = "darkgreen") +
  coord_flip() +
  labs(title = "Top 10 Wicket Takers", x = "Player", y = "Wkts")

sixes <- performance_data[["Most Sixes Per Innings All Seasons Combine"]]

names(sixes) <- make.names(names(sixes))  # X6s becomes X6s

sixes %>%
  group_by(Player) %>%
  summarise(Max_Sixes = max(X6s, na.rm = TRUE)) %>%
  ungroup() %>%
  slice_max(Max_Sixes, n = 10) %>%
  arrange(desc(Max_Sixes)) %>%
  mutate(Player = factor(Player, levels = unique(Player))) %>%
  ggplot(aes(x = Player, y = Max_Sixes)) +
  geom_col(fill = "darkred") +
  coord_flip() +
  labs(title = "Top 10 Most Sixes in a Single Innings", x = "Player", y = "Sixes")
```
Each EDA visualization was designed to support intuitive understanding and drive further questioning.

5. Feature selection
      For this project, we focused on selecting only the features that were actually present and had enough valid data to be useful. We included things like toss decisions, team strengths, venue performance, and home advantage—basically, factors that could realistically impact match outcomes. To keep things clean and avoid errors, we only used features that had enough non-missing values. This approach helped make the models more accurate and kept the code flexible no matter what the dataset looked like.
```{r message=FALSE}
# Defining home venues per team
home_venues <- list(
  "Mumbai Indians" = "Wankhede Stadium",
  "Chennai Super Kings" = "MA Chidambaram Stadium",
  "Royal Challengers Bangalore" = "M Chinnaswamy Stadium",
  "Kolkata Knight Riders" = "Eden Gardens",
  "Delhi Capitals" = "Arun Jaitley Stadium",
  "Sunrisers Hyderabad" = "Rajiv Gandhi International Stadium",
  "Rajasthan Royals" = "Sawai Mansingh Stadium",
  "Punjab Kings" = "Punjab Cricket Association IS Bindra Stadium",
  "Lucknow Super Giants" = "BRSABV Ekana Cricket Stadium",
  "Gujarat Titans" = "Narendra Modi Stadium"
)

home_advantage_df <- enframe(home_venues, name = "Team", value = "home_venue")

# Droping existing conflicting columns if they exist
match_info <- match_info %>%
  select(-any_of(c("team1_home_venue", "team2_home_venue")))

# Joining home venues to match_info
match_info <- match_info %>%
  left_join(home_advantage_df, by = c("team1" = "Team")) %>%
  rename(team1_home_venue = home_venue) %>%
  left_join(home_advantage_df, by = c("team2" = "Team")) %>%
  rename(team2_home_venue = home_venue)

# Creating home advantage flags
match_info <- match_info %>%
  mutate(
    home_advantage_team1 = ifelse(venue == team1_home_venue, 1, 0),
    home_advantage_team2 = ifelse(venue == team2_home_venue, 1, 0)
  )

# Venue strength: win % at each venue
venue_strength <- match_info %>%
  group_by(venue, winner) %>%
  summarise(wins = n(), .groups = "drop") %>%
  rename(Team = winner)

venue_matches <- match_info %>%
  pivot_longer(cols = c(team1, team2), names_to = "team_pos", values_to = "Team") %>%
  group_by(venue, Team) %>%
  summarise(total_played = n(), .groups = "drop")

venue_strength <- left_join(venue_strength, venue_matches, by = c("venue", "Team")) %>%
  mutate(venue_win_pct = ifelse(total_played > 0, wins / total_played, 0))

# Joining venue strength to match_info
match_info <- match_info %>%
  left_join(venue_strength %>% rename(venue_strength_team1 = venue_win_pct), by = c("venue", "team1" = "Team")) %>%
  left_join(venue_strength %>% rename(venue_strength_team2 = venue_win_pct), by = c("venue", "team2" = "Team"))

# Joining team performance stats
match_info <- match_info %>%
  left_join(team_strength %>% select(Team, team1_total_runs = most_runs, team1_total_wickets = most_wickets), by = c("team1" = "Team")) %>%
  left_join(team_strength %>% select(Team, team2_total_runs = most_runs, team2_total_wickets = most_wickets), by = c("team2" = "Team"))

# Finaling model data
model_data <- match_info %>%
  select(venue_strength_team1, venue_strength_team2,
         home_advantage_team1, home_advantage_team2,
         team1_total_runs, team2_total_runs,
         team1_total_wickets, team2_total_wickets,
         toss_decision)

cat("Final features selected for modeling:\n")
print(colnames(model_data))


```
6 : Training and Modeling using random forest and xgboost
        We used two powerful machine learning models to predict match outcomes: Random Forest and XGBoost. Random Forest was tuned using repeated cross-validation and different mtry values to find the best-performing combination. XGBoost was trained with 100 boosting rounds and logistic loss to handle the binary classification task. Once both models were trained, we compared their individual performance and also created ensemble predictions. The ensemble approach combined both models through voting—first simple, then weighted—to boost overall accuracy. This helped us capture the strengths of both algorithms and make more reliable predictions. The final weighted ensemble crossed 70% accuracy, showing a strong improvement from baseline.
```{r}
# Required Libraries
library(caret)
library(randomForest)
library(xgboost)
library(dplyr)
library(stringr)

# Creating binary target
match_info <- match_info %>%
  mutate(
    winner = str_trim(tolower(winner)),
    team1 = str_trim(tolower(team1)),
    team2 = str_trim(tolower(team2)),
    target = case_when(
      winner == team1 ~ "team1",
      winner == team2 ~ "team2",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(target))

# Dynamically adding features only if they exist
good_features <- c()
add_if_available <- function(feature, min_valid = 100) {
  if (feature %in% names(match_info) && sum(!is.na(match_info[[feature]])) >= min_valid) {
    assign("good_features", c(get("good_features", envir = .GlobalEnv), feature), envir = .GlobalEnv)
  }
}

features_to_try <- c(
  "toss_decision", "toss_win", "team1_total_runs", "team2_total_runs",
  "team1_total_wickets", "team2_total_wickets", "venue_strength_team1", 
  "venue_strength_team2", "home_advantage_team1", "home_advantage_team2",
  "head_to_head_wins_team1", "head_to_head_wins_team2"
)

invisible(lapply(features_to_try, add_if_available))
good_features <- c(good_features, "target")

#Preparing model_data
model_data <- match_info %>%
  select(all_of(good_features)) %>%
  filter(complete.cases(.)) %>%
  mutate(target = as.factor(target))

print(table(model_data$target))

# Exit if no class diversity
if (length(unique(model_data$target)) < 2 || nrow(model_data) < 10) {
  stop("Not enough class diversity or data to train model.")
}

# Training/Testing Split
set.seed(123)
train_index <- createDataPartition(model_data$target, p = 0.8, list = FALSE)
train_data <- model_data[train_index, ]
test_data <- model_data[-train_index, ]

# Tuned Random Forest using caret
control <- trainControl(method = "repeatedcv", number = 5, repeats = 3)
tunegrid <- expand.grid(mtry = 2:min(5, ncol(train_data) - 1))

set.seed(123)
rf_tuned <- train(
  target ~ ., data = train_data, method = "rf",
  trControl = control, tuneGrid = tunegrid, ntree = 1000
)

rf_tuned_preds <- predict(rf_tuned, test_data)
cat("Tuned Random Forest Accuracy:\n")
print(mean(rf_tuned_preds == test_data$target))
print(confusionMatrix(rf_tuned_preds, test_data$target))
varImpPlot(rf_tuned$finalModel)

# XGBoost
train_matrix <- model.matrix(target ~ . -1, data = train_data)
train_label <- as.numeric(train_data$target) - 1
test_matrix <- model.matrix(target ~ . -1, data = test_data)
test_label <- as.numeric(test_data$target) - 1

xgb_model <- xgboost(
  data = train_matrix, label = train_label,
  objective = "binary:logistic", nrounds = 100, verbose = 0
)

xgb_preds_prob <- predict(xgb_model, test_matrix)
xgb_preds <- ifelse(xgb_preds_prob > 0.5, "team2", "team1")
cat("XGBoost Accuracy:\n")
print(mean(xgb_preds == test_data$target))

# Ensemble Voting
rf_bin <- as.numeric(rf_tuned_preds == "team1")
xgb_bin <- as.numeric(xgb_preds == "team1")

# Simple Voting
ensemble_preds <- ifelse((rf_bin + xgb_bin) >= 1, "team1", "team2")
cat("Ensemble Accuracy (RF + XGB):\n")
print(mean(ensemble_preds == test_data$target))
print(confusionMatrix(as.factor(ensemble_preds), test_data$target))

# Weighted Voting: RF = 2 votes, XGB = 1 vote
ensemble_preds <- ifelse((2 * rf_bin + xgb_bin) >= 2, "team1", "team2")
cat("Weighted Ensemble Accuracy (2*RF + XGB):\n")
print(mean(ensemble_preds == test_data$target))
print(confusionMatrix(as.factor(ensemble_preds), test_data$target))

```

